\chapter{Introduction}
\begin{shadequote}
The sure and definite determination (of species of bacteria) requires so much time, so much acumen of eye and judgement, so much of perseverance and patience that there is hardly anything else so \mbox{difficult}. \par\emph{Otto F. M\"uller}
\end{shadequote}


\section{Description of Problem}
The quote above, by no mistake, graced the cover of the International Journal of Systematic and Evolutionary Microbiology for decades.
Whereas plant and animal systematists are guided by a theory based approach to \index{demarcating} species, microbiologists have yet to agree on a set of ecological and evolutionary properties that could serve to identify bacterial species~\cite{cohan2007systematics}.
They are naturally handicapped by the paucity of morphological differences that could aide in differentiation of closely related bacterial species.
Another factor is that microbiologists cannot predict which traits will cause a speciation event since bacteria are capable of receiving genes from distant relatives through a process known as horizontal gene transfer (HGT)~\cite{cohan2007systematics}.
Thus, in order to effectively understand the microbiome we must strive towards developing a method for consistently demarcating groups, from bacterial diversity, that play distinct ecological roles~\cite{koeppel2008identifying}.


Initially, closely related bacterial species were identified based on metabolic phenotype.
Systematists now rely on molecular approaches that utilize the decreasing cost of DNA sequencing to compare genetic information.
A 70\% cutoff was established for whole genome hybridization studies (comparing loss and gain of large chunks of DNA), replaced by varying degrees of sequence identities in homologous genes~\cite{cohan2007systematics,carlo}.
From these technological breakthroughs scientists have taken great steps towards understanding bacteria speciation, at the same time they have brought into focus new difficulties.


\subsection{Diversity of Bacterial Species}
Estimates of eukaryotic diversity fall within the range of 10 to 50 million species. Even though we have only observed approximately 9000 prokaryotic species, indirect approaches that do not rely on cultivation hint at the existence of a billion or more prokaryotic species worldwide and 10 million within a given habitat~\cite{cohan2008origins}.
The only rational approach to clustering such a large group of diverse organisms effectively is with a theory based molecular method.


Current protocols for deciding bacterial lineage are functionally incomplete. Recent ecological studies show that a named bacterial species is typically an assemblage of closely related but ecologically distinct populations~\cite{cohan2007systematics}.
Thus, within a named species established by antiquated methods, one may find distantly related organisms that do not naturally fit in with what we define to be a cohesive species cluster.
Our envisioned demarcation algorithm would be capable of identifying putative clusters of ecologically distinct organisms within named bacterial clades.


\subsection{Differences of Bacterial Population Dynamics}
As briefly mentioned earlier their exist peculiarities of bacterial population dynamics that complicate demarcation.

\section{Benefits}%to Understanding Bacteria Speciation or Practical Applications
We must first be able to identify the basic units operating within the system, in order to understand the microbiome.
Once we have established the atomic functional unit we can then start building collections of relationships between these units. 


Bioremediation
Antibiotic resistance
countless health benefits
others

\section{My Thesis Purpose}% SHOULD I PUT THIS SECTION AT THE BEGINNING?

My aims for this project are several. I want to make a production ready version of the optimized Ecotype Simulation (ES2) software.
I want to carry out tests that show the results we get from ES2 are the same as ES1. 
Then I would like to run ES2 in conjunction with other available demarcation programs (ES1, BAPS, GMYC, AdaptML) on generated datasets to show its accuracy and speed optimizations.
Chapter 1 will introduce the problem that ES tries to address, and the benefits that understanding bacterial speciation will achieve.
The following chapter will go over several current molecular models for bacteria speciation, the underlying algorithms that make them function, and describe the design of ES.
Next section discusses our approach to ES optimization, as well as other future enhancements planned.
In the fourth chapter, I will go over comparison results between various demarcation programs.
Finally, I'll conclude with a discussion of what we set out to accomplish compared to what we achieved, shortcomings, and some final thoughts on future directions.

